---
title: "The Shift"
author:
  name: Christian Hicks
  url: https://github.com/hicks017
format:
  html:
    embed-resources: true
    df-print: kable
    toc: true
    toc-depth: 3
    code-fold: true
    date: 2025-05-06
    date-modified: last-modified
description: Exploratory data analysis of the limitations set on defensive shifting.
editor: visual
---

## Setup

```{r}
#| code-summary: Packages
#| message: false

library(readr)
library(dplyr)
library(skimr)
library(ggplot2)
library(broom)
library(gtsummary)
```

```{r}
#| code-summary: Data
#| message: false

# Use a flexible route to the file path
if (basename(getwd()) == "explore data") {
  path <- "../resources/data"
} else if (basename(getwd()) == "MLB-rule-changes") {
  path <- "resources/data"
} else {print("Failed to set path")}

df_babip <- read_csv(file.path(path, "team_babip.csv"))
df_pull <- read_csv(file.path(path, "pull_side_outcomes.csv"))
```

## BABIP

```{r}
#| code-summary: Prepare data

# Change name for Cleveland to the Guardians
df_babip <- mutate(
  df_babip,
  team_name = ifelse(
    endsWith(team_name, "Indians"),
    "Cleveland Guardians",
    team_name
  )
)

# Indicate rule change year
df_babip <- mutate(
  df_babip,
  rule_change = ifelse(season < 2022, FALSE, TRUE),
  season = factor(season)
)
```

Was team BABIP affected after the defensive shift was banned?

```{r}

# Summarize by season
df_babip |>
  select(season, babip) |>
  group_by(season) |>
  skimr::skim() |>
  skimr::yank("numeric")

# Visualize
ggplot(df_babip, aes(x = season, y = babip, fill = rule_change)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 0.4)) +
  theme(
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(angle=0, vjust = 0.5)
  ) +
  ggtitle("Team BABIP by season")

# t test
t.test(babip ~ rule_change, data = df_babip) |>
  tidy() |>
  rename(pre_2022 = estimate1, post_2022 = estimate2)
```

There was a very small decrease (0.004) in team BABIP after the defensive shift was banned.

## Pull side outcomes

```{r}
#| code-summary: Prepare data

df_pull <- mutate(
  df_pull,
  rule_change = ifelse(season < 2022, FALSE, TRUE)
)

```

Were pull side outcomes of at bats affected after the defensive shift was banned?

```{r}

# Summarize by year
gtsummary::tbl_summary(
  data = filter(df_pull, play_outcome != "unknown"), # 3 plays were unknown
  by = rule_change,
  include = play_outcome
) |>
  gtsummary::add_p()
```

There was a very small increase (0.3%) in hits after the defensive shift ban, that was accompanied by a small decrease (0.2%) in errors. Out occurrence saw the least change.

Note: A limitation is that there is no simple way to fetch data on plays where the defensive shift formation was indicated. If that data existed then we could remove plays from this analysis where the defensive shift was not used. Therefore, the results here may be affected by the noise of plays that did not have a defensive shift involved.
