---
title: "Extra-innings Baserunner"
author:
  name: Christian Hicks
  url: https://github.com/hicks017
format:
  html:
    embed-resources: true
    df-print: kable
    toc: true
    toc-depth: 3
    code-fold: true
    date: 2025-05-03
    date-modified: last-modified
description: Exploratory data analysis of the extra-innings baserunner on second base.
editor: visual
---

## Setup

```{r}
#| code-summary: Packages
#| message: false

library(readr)
library(dplyr)
library(skimr)
library(ggplot2)
library(broom)
```

```{r}
#| code-summary: Data
#| message: false

# Use a flexible route to the file path
if (basename(getwd()) == "explore data") {
  path <- "../resources/data"
} else if (basename(getwd()) == "MLB-rule-changes") {
  path <- "resources/data"
} else {print("Failed to set path")}

df_extras <- read_csv(file.path(path, "extra_innings_games.csv"))
```

## Total extra innings

```{r}
#| code-summary: Prepare data

# Change name for Cleveland to the Guardians
df_extras <- mutate(
  df_extras,
  team_winner = ifelse(
    endsWith(team_winner, "Indians"),
    "Cleveland Guardians",
    team_winner
  ),
  team_loser = ifelse(
    endsWith(team_loser, "Indians"),
    "Cleveland Guardians",
    team_loser
  )
)

# Count innings over 9 per game per year
df_extras <- 
  df_extras |>
  mutate(extras = final_inning - 9) |>
  
  # Add indication of rule change and change year to a factor
  mutate(rule_change = ifelse(year < 2020, FALSE, TRUE)) |>
  mutate(year = factor(year))
```

How many extra innings were required to complete a game over the years?

```{r}

# Summarize by year
df_extras |>
  select(year, extras) |>
  group_by(year) |>
  skimr::skim() |>
  skimr::yank("numeric")

# Visualize
ggplot(df_extras, aes(x = year, y = extras, fill = rule_change)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_y_continuous(breaks = seq(0, 10, by = 1)) +
  theme(
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(angle=0, vjust = 0.5)
  ) +
  ggtitle("Total extra innings per game by year")

# t test
t.test(extras ~ rule_change, data = df_extras) |>
  tidy() |>
  rename(pre_2020 = estimate1, post_2020 = estimate2)
```

Number of extra innings required to complete a game decreased by just under 1 inning after the rule change.

Note that outlying "marathon" games are much shorter after the rule change, dropping from 7-10 extra innings to 4-5.

## Total runs in extra innings

How many runs were scored during extra innings?

```{r}

# Summarize by year
df_extras |>
  select(year, total_extra_runs) |>
  group_by(year) |>
  skimr::skim() |>
  skimr::yank("numeric")

# Visualize
ggplot(df_extras, aes(x = year, y = total_extra_runs, fill = rule_change)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_y_continuous(breaks = seq(0, 12, by = 2)) +
  theme(
    axis.title.y = element_text(angle=0, vjust = 0.5)
  ) +
  ggtitle("Total runs scored during extra innings per game by year")

# t test
t.test(total_extra_runs ~ rule_change, data = df_extras) |>
  tidy() |>
  rename(pre_2020 = estimate1, post_2020 = estimate2)
```

Number of runs scored during extra innings per game game increased by almost 1 run after the rule change. The median increased by 2 runs.

## Batters until first score

How many batters appear before the first run is scored in extra innings?

```{r}
# Summarize by year
df_extras |>
  select(year, batters_until_first_run) |>
  group_by(year) |>
  skimr::skim() |>
  skimr::yank("numeric")

# Visualize
ggplot(df_extras, aes(x = year, y = batters_until_first_run, fill = rule_change)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme(axis.title.y = element_text(angle=0, vjust = 0.5)) +
  ggtitle("Batters until first run scored in extras per game by year")

# t test
t.test(batters_until_first_run ~ rule_change, data = df_extras) |>
  tidy() |>
  rename(pre_2020 = estimate1, post_2020 = estimate2)
```

Number of batters until first run scored during extra innings per game decreased by more than 8 after the rule change.

Note that the 25th percentile dropped from 5-6 to 2 batters, and the 75th percentile dropped from 16-20 to 7. Additionally, outlying games dropped from 40-60 to 13-30 These are quite big changes.
