---
title: "Durations of plays and games"
author:
  name: Christian Hicks
  url: https://github.com/hicks017
format:
  html:
    embed-resources: true
    df-print: kable
    toc: true
    toc-depth: 3
    code-fold: true
    date: 2025-05-04
    date-modified: last-modified
description: Exploratory data analysis of the durations of play.
editor: visual
---

## Setup

```{r}
#| code-summary: Packages
#| message: false

library(readr)
library(dplyr)
library(skimr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(broom)
```

```{r}
#| code-summary: Data
#| message: false

# Use a flexible route to the file path
if (basename(getwd()) == "explore data") {
  path <- "../resources/data"
} else if (basename(getwd()) == "MLB-rule-changes") {
  path <- "resources/data"
} else {print("Failed to set path")}

df_duration <- read_csv(file.path(path, "play_duration.csv"))

# Factorize season
df_duration <- mutate(df_duration, season = factor(season))
```

## Individual play duration

Has there been any change in the length of individual plays since the 2023 rule changes?

Note: Specifically pitch clock. Reduced pitcher disengagements could play a role too.

### Data preparation

```{r}
# Summarize by season
df_duration |>
  select(season, duration_sec) |>
  group_by(season) |>
  skimr::skim() |>
  skimr::yank("numeric")
```

Many seasons recorded plays that had negative values and extremely high values. These major outliers need to be removed before performing any analyses.

```{r}
old_scipen <- getOption("scipen")
options(scipen = 999)

# Define the percentiles
probs <- c(
  1,
  0.9999,
  0.999,
  0.99,
  0.95,
  0.9,
  0.05,
  0.01,
  0.001,
  0.0001,
  0
)

# Summarize various percentiles of play duration
df_duration |>
  summarise(quantiles = list(quantile(duration_sec, probs = probs))) |>
  unnest_wider(quantiles) |>
  pivot_longer(everything(), names_to = "Percentile", values_to = "Seconds")

options(scipen = old_scipen)  # restore original value

# Visualize with cumaltive distribution (filtered)
df_duration |>
  filter(duration_sec > 0 & duration_sec < 423) |>
  ggplot(aes(duration_sec)) +
  stat_ecdf(geom = "step") +
  labs(
    title = "Cumulative density function for play duration",
    caption = "Filtered for 0 < duration_sec < 423 (99.9%ile)"
  )

# Visualize with histogram (filtered)
tbl_duration_filtered_pre <- filter(
  df_duration,
  duration_sec > 0
    & duration_sec < 271 # 95th percentile
    & year(startTime) < 2023
)
tbl_duration_filtered_post <- filter(
  df_duration,
  duration_sec > 0
    & duration_sec < 271 # 95th percentile
    & year(startTime) >= 2023
)

tbl_duration_filtered_pre$group <- "Pre rule change"
tbl_duration_filtered_post$group <- "Post rule change"

ggplot() +
  
  # Post rule change
  geom_density(
    data = tbl_duration_filtered_post,
    aes(x = duration_sec, fill = group),
    color = NA
  ) +
  
  # Pre rule change
  geom_density(
    data = tbl_duration_filtered_pre,
    aes(x = duration_sec, fill = group),
    color = NA
  ) +
  
  # Post rule change (line)
  geom_density(
    data = tbl_duration_filtered_post,
    aes(x = duration_sec),
    color = "orange",
    linewidth = 2
  ) +
  
  # Pre rule change (line)
  geom_density(
    data = tbl_duration_filtered_pre,
    aes(x = duration_sec),
    color = "black"
  ) +
  
  # Formatting
  scale_fill_manual(
    name = "Rule Change",
    values = c("Pre rule change" = "skyblue", "Post rule change" = "orange")
  ) +
  
  scale_x_continuous(breaks = seq(0, 250, by = 50), expand = c(0,0)) +
  
  theme(
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(angle = 0, vjust = 0.5),
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.8),
    legend.title = element_blank()
  ) +
  
  labs(
    title = "Shorter play duration after pitch clock implementation",
    x = "Duration (seconds)",
    y = "Density\nof plays"
  )
```

### Analysis

```{r}
# Indicate rule change
df_duration <- mutate(
  df_duration,
  rule_change = ifelse(year(startTime) < 2023, FALSE, TRUE)
)

# t test
t.test(duration_sec ~ rule_change, data = df_duration) |>
  tidy() |>
  rename(pre_2023 = estimate1, post_2023 = estimate2)
```

Mean play duration decreased by 48 seconds after the implementation of the pitch clock and the pitcher disengagement policy.
