---
title: "Three batter minimum"
author:
  name: Christian Hicks
  url: https://github.com/hicks017
format:
  html:
    embed-resources: true
    df-print: kable
    toc: true
    toc-depth: 3
    code-fold: true
    date: 2025-04-29
    date-modified: last-modified
description: Exploratory data analysis of the three batter minimum.
editor: visual
---

## Setup

```{r}
#| code-summary: Packages
#| message: false

library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(skimr)
library(gtsummary)
```

```{r}
#| code-summary: Data
#| message: false

# Use a flexible route to the file path
if (basename(getwd()) == "explore data") {
  path <- "../resources/data"
} else if (basename(getwd()) == "MLB-rule-changes") {
  path <- "resources/data"
} else {print("Failed to set path")}

df_appearances <- read_csv(file.path(path, "pitcher_appearances.csv"))
df_changes <- read_csv(file.path(path, "pitcher_changes.csv"))
```

## Pitching appearances

```{r}
#| code-summary: Prepare data

# Remove duplicates
df_appearances <- distinct(df_appearances)

# Change name for Cleveland to the Guardians
df_appearances <- mutate(df_appearances, team_name = ifelse(
  endsWith(team_name, "Indians"),
  "Cleveland Guardians",
  team_name
))

# Count pitching changes per game by year
tbl_n_per_game_yearly <- 
  df_appearances |>
  group_by(game_id, year) |>
  summarise(n = n(), .groups = "drop") |>
  
  # Add indication of rule change and change year to a factor
  mutate(rule_change = ifelse(year < 2020, FALSE, TRUE)) |>
  mutate(year = factor(year))
```

How many pitchers are teams using in a game over the years?

With a three-batter minimum, the LOOGY and similar specialty roles should mostly be eliminated leading to fewer pitchers used in a game.

```{r}
# Plot pitchers per game per year
ggplot(tbl_n_per_game_yearly, aes(x = year, y = n, fill = rule_change)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme(axis.title.y = element_text(angle=0, vjust = 0.5)) +
  ggtitle("Total pitchers per game by year")

# Descriptive statistics
skimr::skim(group_by(select(tbl_n_per_game_yearly, -game_id), year)) |> 
  skimr::yank("numeric")

# t test
t.test(n ~ rule_change, data = tbl_n_per_game_yearly) |>
  tidy() |>
  rename(pre_2020 = estimate1, post_2020 = estimate2)
```

Number of pitchers used by both teams has increased since 2020, but the magnitude was small.

## Mid-inning pitching changes

How many pitching changes occurred during inning play in a game over the years?

```{r}
#| code-summary: Prepare data

# Change name for Cleveland to the Guardians
df_changes <- mutate(df_changes, team_name = ifelse(
  endsWith(team_name, "Indians"),
  "Cleveland Guardians",
  team_name
))

# Count pitching changes per game by year
tbl_changes_per_game_yearly <- 
  df_changes |>
  group_by(game_id, year) |>
  summarise(n = n(), .groups = "drop") |>
  
  # Input games without a mid-inning pitching change
  mutate(event = TRUE) |>
  right_join(
    df_appearances |> select(game_id, year) |> distinct(),
    by = c("game_id", "year")
  ) |>
  mutate(n = ifelse(is.na(event), 0, n)) |>
  select(-event) |>
  
  # Add indication of rule change and change year to a factor
  mutate(rule_change = ifelse(year < 2020, FALSE, TRUE)) |>
  mutate(year = factor(year))
```

A three-batter minimum means there's less opportunity to call in a pitcher for a single batter, which should decrease the number of pitching changes that occur in the middle of an inning.

```{r}
# Plot pitchers per game per year
ggplot(tbl_changes_per_game_yearly, aes(x = year, y = n, fill = rule_change)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme(axis.title.y = element_text(angle=0, vjust = 0.5)) +
  ggtitle("Total mid-inning pitching changes per game by year")

# Descriptive statistics
skimr::skim(group_by(select(tbl_changes_per_game_yearly, -game_id), year)) |> 
  skimr::yank("numeric")

# t test
t.test(n ~ rule_change, data = tbl_changes_per_game_yearly) |>
  tidy() |>
  rename(pre_2020 = estimate1, post_2020 = estimate2)
```

The number of mid-inning pitching changes decreased since 2020, but the magnitude was small.

## Games with unscheduled pauses

A mid-inning pitching change would add a pause in play that was not already scheduled, such as pauses in between innings.

What proportion of games had these added pauses?

### 1 or more pauses

```{r}
#| code-summary: Prepare data

# Indicate games with n pauses
tbl_changes_per_game_yearly <- mutate(
  tbl_changes_per_game_yearly,
  pause_01 = ifelse(n > 0, TRUE, FALSE),
  pause_02 = ifelse(n > 1, TRUE, FALSE),
  pause_03 = ifelse(n > 2, TRUE, FALSE)
)

# Summarize by year
skimr::skim(select(group_by(tbl_changes_per_game_yearly, year), pause_01, year)) |>
  skimr::yank("logical")

# Test result by rule change
gtsummary::tbl_summary(
  data = tbl_changes_per_game_yearly,
  include = pause_01,
  by = rule_change
) |>
  gtsummary::  modify_spanning_header(all_stat_cols() ~ "**Rule change**") |>
  gtsummary::add_p()
```

There was a -1% change in games with any added pauses in play due to mid-inning pitching changes.

### 2 or more pauses

```{r}
# Summarize by year
skimr::skim(select(group_by(tbl_changes_per_game_yearly, year), pause_02, year)) |>
  skimr::yank("logical")

# Test result by rule change
gtsummary::tbl_summary(
  data = tbl_changes_per_game_yearly,
  include = pause_02,
  by = rule_change
) |>
  gtsummary::  modify_spanning_header(all_stat_cols() ~ "**Rule change**") |>
  gtsummary::add_p()
```

There was a -3% change in games with 2 or more added pauses in play due to mid-inning pitching changes.

### 3 or more pauses

```{r}
# Summarize by year
skimr::skim(select(group_by(tbl_changes_per_game_yearly, year), pause_03, year)) |>
  skimr::yank("logical")

# Test result by rule change
gtsummary::tbl_summary(
  data = tbl_changes_per_game_yearly,
  include = pause_03,
  by = rule_change
) |>
  gtsummary::  modify_spanning_header(all_stat_cols() ~ "**Rule change**") |>
  gtsummary::add_p()
```

There was a -4% change in games with 3 or more added pauses in play due to mid-inning pitching changes.

**Overall conclusion:** Although reductions of games with added pauses are statistically significant and of greater magnitude as the minimum threshold increases, the real-world impact is likely small.

Note: Could just show `pause_01` through `pause_03` on same table.
