---
title: "Three batter minimum"
author:
  name: Christian Hicks
  url: https://github.com/hicks017
format:
  html:
    embed-resources: true
    df-print: kable
    toc: true
    toc-depth: 3
    code-fold: true
    date: 2025-04-29
    date-modified: last-modified
description: Exploratory data analysis of the three batter minimum.
editor: visual
---

## Setup

```{r}
#| code-summary: Packages
#| message: false

library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(skimr)
```

```{r}
#| code-summary: Data
#| message: false

# Use a flexible route to the file path
if (basename(getwd()) == "explore data") {
  path <- "../resources/data/"
} else if (basename(getwd()) == "MLB-rule-changes") {
  path <- "resources/data/"
} else {print("Failed to set path")}

df_appearances <- read_csv(file.path(path, "pitcher_appearances.csv"))
df_changes <- read_csv(file.path(path, "pitcher_changes.csv"))
```

## Pitching appearances

```{r}
#| code-summary: Prepare data

# Change name for Cleveland to the Guardians
df_appearances <- mutate(df_appearances, team_name = ifelse(
  endsWith(team_name, "Indians"),
  "Cleveland Guardians",
  team_name
))

# Count pitching changes per game by year
tbl_n_per_game_yearly <- 
  df_appearances |>
  group_by(game_id, year) |>
  summarise(n = n(), .groups = "drop") |>
  
  # Add indication of rule change and change year to a factor
  mutate(rule_change = ifelse(year < 2020, FALSE, TRUE)) |>
  mutate(year = factor(year))
```

How many pitchers are teams using in a game over the years?

With a three-batter minimum, the LOOGY specialty role should mostly be eliminated and fewer pitchers might be used in a game.

```{r}
# Plot pitchers per game per year
ggplot(tbl_n_per_game_yearly, aes(x = year, y = n, fill = rule_change)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme(axis.title.y = element_text(angle=0, vjust = 0.5)) +
  ggtitle("Total pitchers per game by year")

# Descriptive statistics
skimr::skim(group_by(select(tbl_n_per_game_yearly, -game_id), year)) |> 
  skimr::yank("numeric")

# t test
t.test(n ~ rule_change, data = tbl_n_per_game_yearly) |>
  tidy() |>
  rename(pre_2020 = estimate1, post_2020 = estimate2)
```

Number of pitchers used by both teams has increased since 2020, but the magnitude was small.

## Mid-inning pitching changes

How many pitching changes occurred during inning play in a game over the years?

```{r}
#| code-summary: Prepare data

# Change name for Cleveland to the Guardians
df_changes <- mutate(df_changes, team_name = ifelse(
  endsWith(team_name, "Indians"),
  "Cleveland Guardians",
  team_name
))

# Count pitching changes per game by year
tbl_changes_per_game_yearly <- 
  df_changes |>
  group_by(game_id, year) |>
  summarise(n = n(), .groups = "drop") |>
  
  # Add indication of rule change and change year to a factor
  mutate(rule_change = ifelse(year < 2020, FALSE, TRUE)) |>
  mutate(year = factor(year))
```

A three-batter minimum means there's less opportunity to call in a pitcher for a single batter, which should decrease the number of pitching changes that occur in the middle of an inning.

```{r}
# Plot pitchers per game per year
ggplot(tbl_changes_per_game_yearly, aes(x = year, y = n, fill = rule_change)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme(axis.title.y = element_text(angle=0, vjust = 0.5)) +
  ggtitle("Total mid-inning pitching changes per game by year")

# Descriptive statistics
skimr::skim(group_by(select(tbl_changes_per_game_yearly, -game_id), year)) |> 
  skimr::yank("numeric")

# t test
t.test(n ~ rule_change, data = tbl_changes_per_game_yearly) |>
  tidy() |>
  rename(pre_2020 = estimate1, post_2020 = estimate2)
```

The number of mid-inning pitching changes decreased since 2020, but the magnitude was small.
